---
title: "Accessing Data Using STAC API and R"
output:
  html_document:
    df_print: paged
---

### Introduction

This notebook demonstrates how to use the STAC API with the [`` `rstac` `` R package](https://brazil-data-cube.github.io/rstac/) to retrieve elevation data. The demo assumes some familiarity with R.

STAC (SpatioTemporal Asset Catalogue) aims to standardise the way geospatial assets (like satellite imagery, aerial photos, or any other geospatial data) are described and catalogued, making it easier for users to search, discover, and use this data.

STAC APIs provide access to a set of data that users can query and retrieve. You can find a list of STAC APIs at the [STAC Index](https://stacindex.org/) or [STAC Browser](https://radiantearth.github.io/stac-browser/#/?.language=en). Here, we will use [Microsoft Planetary Computer](https://planetarycomputer.microsoft.com/) STAC, which provides 123 collections of open datasets so far.

All necessary packages are available from CRAN and can be installed using `install.packages(c("rstac", "sf", "tidyverse", "terra", "tidyterra", "leaflet", "osmdata"))` .

```{r}
# list of packages
packages <- c("rstac", "sf", "tidyverse", "terra", "tidyterra", "leaflet", "osmdata")
# Install packages that are not installed
install.packages(setdiff(packages, rownames(installed.packages())))
```

### Load libraries

```{r message=TRUE, warning=TRUE, include=FALSE}
# Load necessary libraries
library(rstac)
library(sf)
library(tidyverse)
library(terra)
library(tidyterra)
library(leaflet)
library(osmdata)
```

### Get basin boundary from hydrobasins

Before we start fetching data from the Microsoft Planetary Computer via the STAC API, let's get a catchment boundary to demonstrate how to retrieve data at a catchment level. For this, we use the [HydroBASINS dataset](https://www.hydrosheds.org/products/hydrobasins), which provides polygon layers of sub-basin boundaries at a global scale. The code below downloads the basins for Africa, define a location and extracts the catchment at a chosen location. Here, I am using level 4 data.

```{r}
# Download and Unzip Shapefile
# URL of the zip file
zip_file_url <- "https://data.hydrosheds.org/file/HydroBASINS/standard/hybas_af_lev04_v1c.zip"
# Destination directory and file path
data_directory <- "./02.data" 
dir.create(data_directory) 
zip_file_path <- paste0(data_directory, '/catch_boundary.zip')
# Download the zip file
download.file(zip_file_url, zip_file_path, mode = "wb")
# Unzip the file
unzip(zip_file_path, exdir = data_directory)
# Remove the zip file after unzipping
file.remove(zip_file_path)
```

Next, we choose a location and perform a spatial intersection to find the catchment level 4 polygon at that location. You can change the location to any place across Africa. I chose the Sebou River basin, which is the second largest basin in Morocco, covering approximately 40,000 sq km. This basin is important for its agricultural productivity.

```{r}
# Find the shapefile in the directory
shapefile_path <- list.files(data_directory, pattern = 'lev04.*\\.shp$', full.names = TRUE)
# Read the shapefile
basin_shape_data <- sf::st_read(shapefile_path[1])
# Create an sf object for the Sebou basin location
sebou_location <- st_point(c(-5.93, 34.18)) # change here for your basin
# simple feature column
sebou_location_sf <- st_sfc(sebou_location, crs = st_crs(basin_shape_data))
# Validate and correct the geometries
basin_shape_data <- st_make_valid(basin_shape_data)
```

get the basin and the bounding box

```{r}
# Perform spatial intersection to find the polygon containing the location
intersection <- st_intersects(basin_shape_data, sebou_location_sf, sparse = FALSE)
selected_basin <- basin_shape_data[which(intersection), ]
# Get the bounding box of the selected basin
basin_bbox <- st_bbox(selected_basin)
```

### River network

We obtain the river network (the main Sebou River) from OSM using the `osmdata` package. Note that OSM river network data are not always accurate, so you may want to choose other data sources if the purpose is not purely for visualisation.

```{r}
# Get Main River Data from OSM
# Query OSM data for the Sebou River within the basin bounding box
sebou_river_net <- opq(bbox = basin_bbox, timeout = 300) %>%
  add_osm_feature(key = 'waterway', value = 'river') %>%
  add_osm_feature(key = 'name:en', value = 'Sebou River') %>%
  osmdata_sf()

```

### Interactive map

Now let's add all the layers to a Leaflet map.

```{r}
# Create a Leaflet map to visualise the basin and river
leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addPolygons(data = selected_basin, color = "blue", weight = 2, fillOpacity = 0.5) %>%
  addPolylines(data = sebou_river_net$osm_lines, color = "blue", weight = 2)

```

------------------------------------------------------------------------

### Microsoft Planetary Computer STAC API

#### Example 1: NASADEM Elevation data

We use [Microsoft’s Planetary Computer](https://planetarycomputer.microsoft.com/) to get elevation data over our catchment. There are many elevation products, and here we demonstrate how to get the NASADEM product.

First, we need to let `rstac` know which STAC API we want to query and download from. We do this by passing the URL of the Planetary Computer STAC API to the `rstac::stac()` function.

```{r}
# STAC API Interaction
# Define the URL of the STAC API endpoint
stac_api_url <- "https://planetarycomputer.microsoft.com/api/stac/v1/"
# Create a STAC client instance
stac_client <- stac(stac_api_url) 
# get all MPC collections
stac_client %>% 
  rstac::collections() %>% 
  rstac::get_request()
```

Let's select the *nasadem* elevation.

```{r}
# Collection name for NASADEM data
nasadem_coll <- "nasadem"

```

Now we use `rstac::stac_search()` to pass the collection name and query what items are available in this collection. We will only get items that intersect with our area of interest.

Note the authentication process is built into `rstac`, using the `items_sign()` and `sign_planetary_computer()` functions. It handles the necessary authentication by signing the requests, allowing access to the data.

```{r}
# Search for the collection that intersects with the selected basin
nasadem_data <- stac_client %>%
  stac_search(collections = nasadem_coll,
              intersects = selected_basin) %>%
  rstac::post_request() %>%
  items_sign(sign_fn = sign_planetary_computer())
nasadem_data
```

Let's explore some `rstac` functions to further investigate our collection.

```{r}
# Get the length of the collection
nasadem_data %>% items_length()
# Convert items to a tibble for easier handling
nasadem_items_df <- nasadem_data %>%
  items_as_tibble()
# Retrieve properties and assets for further analysis
nasadem_data %>% items_properties()
nasadem_data %>% items_assets()
```

We get the first item and use the `rast()` function from the `terra` package to render and plot it. We then get the URL (`href`) that points to the location of the data.

```{r}
# Plot Elevation Data
# Extract the URL of the first item's elevation data
first_item <- nasadem_data$features[[1]]
first_item_url <- first_item$assets$elevation$href
# Plot the first elevation tile
terra::rast(first_item_url) %>%
  terra::plot()
```

Next, we get all tile `href` URLs using `map_chr()` from the `purrr` package, then read all files by mapping the `rast` function. The last step is to mosaic and crop the raster to the basin's shape.

```{r}
# Extract elevation URLs from all features
elevation_urls <- map_chr(nasadem_data$features, ~ .x$assets$elevation$href)
# Read and mosaic the rasters
mosaic_raster <- map(elevation_urls, terra::rast) %>%
  terra::sprc() %>% #create spatial collection 
  terra::mosaic()
```

Crop the raster

```{r}
# Crop and mask the mosaic raster with the selected basin
mosaic_raster <- mosaic_raster %>%
  terra::crop(selected_basin) %>% 
  terra::mask(selected_basin)
# Simple plot of the mosaic raster
terra::plot(mosaic_raster)
```

Finally, create a `ggplot` visualisation of the elevation data.

```{r}
# Create a ggplot visualization of the elevation data
ggplot() +
  geom_spatraster(data = mosaic_raster) +
  scale_fill_whitebox_c(palette = "muted", labels = scales::label_number(), n.breaks = 12, guide = guide_legend(reverse = TRUE)) +
  geom_sf(data = selected_basin, fill = "transparent", color = "black", linewidth = 0.5) +
  geom_sf(data = sebou_river_net$osm_lines, color = "blue", linewidth = 0.5) +
  labs(fill = "Elevation [m]") +
  theme_bw()
```

#### Example 2: CMIP6 Climate projections

[Earth Exchange Global Daily Downscaled Projections](https://planetarycomputer.microsoft.com/dataset/nasa-nex-gddp-cmip6) dataset provides global downscaled climate scenarios from CMIP6 GCM runs, supporting the IPCC AR6 report. It includes high-resolution, bias-corrected projections for two Tier 1 SSP emissions scenarios, useful for assessing climate change impacts sensitive to fine-scale climate gradients and local topography.

```{r}
# Define collection and datetime range
collection <- "nasa-nex-gddp-cmip6"
datetime <- "2050-01-01/2060-12-31"

# Search the STAC catalog
data_collection <- stac_client %>%
  stac_search(collections = collection,
              intersects = selected_basin,
              datetime = datetime) %>%
  ext_query("cmip6:model" == 'ACCESS-CM2') %>%
  ext_query("cmip6:scenario" == 'ssp585') %>%
  post_request() %>%
  items_sign(sign_fn = sign_planetary_computer())
```

select temperature asset, get first item and download to disk

```{r}
# Get temperature and first item
tas_item <- data_collection %>% 
  assets_select(asset_names = "tas")
# Get the first item 
first_item <- tas_item$features[[1]]
# Download the asset
first_item %>% 
  assets_download(overwrite = TRUE)
```

List the file and use `rast()` from `terra` to read nc file

```{r}
# Define path and list files
nc_path <- paste0(getwd(),'/nex-gddp-cmip6/NEX/GDDP-CMIP6/ACCESS-CM2/ssp585/r1i1p1f1/tas/')
tas_files <- list.files(nc_path, full.names = TRUE)

# Load the raster and process it
tas_nc <- terra::rast(tas_files)
r_day1 <- tas_nc[[1]] %>% terra::rotate()

# Crop and mask the raster to the basin
tas_basin <- r_day1 %>% 
  terra::crop(selected_basin) %>% 
  terra::mask(selected_basin)
```

quick plot:

```{r warning=FALSE}
# Plot the raster and basin
tas_time <- as.character(terra::time(tas_basin))
terra::plot(tas_basin, main = paste0('tas: ', tas_time), col = rainbow(100))
terra::plot(selected_basin, add = TRUE, border = "black", col = NA)
```

You can convert terra raster to a data frame and get the average temperature across the basin for example

```{r}
# Convert raster to data frame and calculate mean temperature in Celsius
tas_df <- terra::as.data.frame(tas_basin)
mean_tas_celsius <- mean(tas_df$tas_1 - 273.15, na.rm = TRUE)
print(mean_tas_celsius)
```

add to the leaflet map

```{r}
# leaflet 
leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addPolygons(data = selected_basin, weight = 2, fill = FALSE) %>%
  addPolylines(data = sebou_river_net$osm_lines, weight = 2) %>%
  addRasterImage(tas_basin, colors = colorNumeric(rainbow(100),values(tas_basin), na.color = "transparent")) %>%
  addLegend(pal = colorNumeric(rainbow(100),values(tas_basin), na.color = "transparent"),values = values(tas_basin), title = "Temperature (K)")
```
